Algorithmic detection of conserved quantities for finite-difference schemes\thanks{Supported by King Abdullah University of Science and Technology (KAUST) baseline funds and KAUST OSR-CRG2021-4674.}

### 2023-05-09 Papers/Paper CASC format/tmp/CPpaper.tex
...

\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage[latin1]{inputenc}
\usepackage{hyperref}
\usepackage{dsfont}
\usepackage[linesnumbered,ruled,vlined]{algorithm2e}
\usepackage{ifthen}
\usepackage{mmacells}

@@

Here, all quantities, whose conservation in time we check, are functions $\mathcal{J}\in\mathcal{F}(\mathcal{P}(\Ii\times\Nn^l,\Rr^m),\mathcal{D}(\Nn^l))$. 
Let $\tilde{n}:=(n_1,\dots,n_d,s_1,\dots,s_l)\in\Ii\times\Nn^l$. Using the Kronecker delta $\delta$, we rewrite
\begin{gather*}
	\mathcal{J}[u](t_1,\ldots,t_{l})
	=\sum_{\tilde{n}\in\Ii\times\Nn^l}F_{\tilde{n}}[u]\delta_{(t_1,\ldots,t_{l})}(s_1,\ldots,s_l).
\end{gather*}
Let $u,v\in\mathcal{P}(\Ii\times\Nn^l,\Rr^m)$, then
\footnotesize
\begin{gather*}
	D^l\mathcal{J}[u](v)(t_1,\ldots,t_{l}):=\frac{d}{d\epsilon}\mathcal{J}[u+\epsilon v]\Big|_{\epsilon=0}(t_1,\ldots,t_{l})\\
	=\sum_{\tilde{n}}D_1G(u(\tilde{n}+e^1),\ldots,u(\tilde{n}+e^k))\delta_{(t_1,\ldots,t_{l})}(s_1,\ldots,s_l)v(\tilde{n}+e^1)\\
	+D_2G(u(\tilde{n}+e^1),\ldots,u(\tilde{n}+e^k))\delta_{(t_1,\ldots,t_{l})}(s_1,\ldots,s_l)v(\tilde{n}+e^2)\\+\ldots
	+D_kG(u(\tilde{n}+e^1),\ldots,u(\tilde{n}+e^k))\delta_{(t_1,\ldots,t_{l})}(s_1,\ldots,s_l)v(\tilde{n}+e^k)\\
	=\sum_{\tilde{n}}D_1G(u(\tilde{n}),\ldots,u(\tilde{n}+e^k-e^1))\delta_{(t_1,\ldots,t_{l})}(s_1-e^1_{d+1},\ldots,s_l-e^1_{d+l})v(\tilde{n})\\
	+D_2G(u(\tilde{n}+e^1-e^2),\ldots,u(\tilde{n}+e^k-e^2))\delta_{(t_1,\ldots,t_{l})}(s_1-e^2_{d+1},\ldots,s_l-e^2_{d+l})v(\tilde{n})\\+\ldots
	+D_kG(u(\tilde{n}+e^1-e^k),\ldots,u(\tilde{n}))\delta_{(t_1,\ldots,t_{l})}(s_1-e^k_{d+1},\ldots,s_l-e^k_{d+l})v(\tilde{n})\\
	=\sum_{\tilde{n}}\Big(D_1G(u(\tilde{n}),\ldots,u(\tilde{n}+e^k-e^1))\delta_{(t_1+e^1_{d+1},\ldots,t_{l}+e^1_{d+l})}(s_1,\ldots,s_l)\\
	+D_2G(u(\tilde{n}+e^1-e^2),\ldots,u(\tilde{n}+e^k-e^2))\delta_{(t_1+e^2_{d+1},\ldots,t_{l}+e^2_{d+l})}(s_1,\ldots,s_l)\\+\ldots
	+D_kG(u(\tilde{n}+e^1-e^k),\ldots,u(\tilde{n}))\delta_{(t_1+e^k_{d+1},\ldots,t_{l}+e^k_{d+l})}(s_1,\ldots,s_l)\Big)v(\tilde{n})\\
	=:\sum_{\tilde{n}}\mathcal{V}^l(\mathcal{J})[u](\tilde{n},t)v(\tilde{n}).
\end{gather*}
\normalsize
\begin{definition}
$\mathcal{V}^l(\mathcal{J})$ is the \emph{discrete partial variational derivative} of $\mathcal{J}$.
	Note that $\mathcal{V}^0=\mathcal{V}$.
\end{definition}
\begin{example}
	\label{dpvd1}
	The discrete partial variational derivative of 
	\begin{gather*}
		\mathcal{J}[u](t)=\sum_nu(n,t+1)-u(n,t)=\sum_{n,s}(u(n,s+1)-u(n,s))\delta_t(s)
	\end{gather*}
	is $\delta_{t+1}(s)-\delta_t(s)$, because
	\small
	\begin{gather*}
		D^1(\mathcal{J})[u](v)(t)\\
		=\frac{d}{d\epsilon} \sum_{n,s}((u(n,s+1)+\epsilon v(n,s+1))-(u(n,s)+\epsilon v(n,s)))\delta_t(s)\\
		=\sum_{n,s}v(n,s+1)\delta_t(s)-v(n,s)\delta_t(s)=\sum_{n,s}v(n,s)\delta_t(s-1)-v(n,s)\delta_t(s)\\
		=\sum_{n,s}(\delta_{t+1}(s)-\delta_t(s))v(n,s)=\sum_{n,s}\mathcal{V}^1(\mathcal{J})(n,s,t)v(n,s).
	\end{gather*}
	\normalsize
\end{example}
\begin{example}
	In our {\sc Mathematica} implementation, Example \ref{dpvd1} is computed by

	\begin{small}
		\begin{mmaCell}[moredefined={variables, expression, PartialDVarD}]{Input}
  variables=<|"indVars"\(\pmb{\to}\)\{n,t\},"depVars"\(\pmb{\to}\)\{u\},
  "timeVars"\(\pmb{\to}\)\{t\}|>
  PartialDVarD[variables][u[n,t+1]-u[n,t]]

@@

	\caption{{\sc time-explicit DiscreteConservedQ}}
	\label{alg: discreteconservedqexplicit}
\end{algorithm}
The result of Step 2 exists only at time $t$ because all instances of $t+1$ have been replaced. Thus, we do not need any computations of difference ideals or Gr{\"o}bner basis.
\begin{example}
	\label{timeexplicitheateqalgo}
	We examine the conservation of $\sum_nu(n,t)$ under the following scheme.
	\begin{equation*}
		u(n,t+1)-u(n,t)-(u(n+1,t)-2u(n,t)+u(n-1,t)).
	\end{equation*}
	The discrete time derivative (Step 1) is
	\begin{gather*}
		\sum_{n\in\Ii}u(n,t+1)-u(n,t).
	\end{gather*}
	Replace $u(n,t+1)$ by the right-hand side (Step 2) to get
	\begin{gather*}
		\sum_{n\in\Ii}u(n,t)+(u(n+1,t)-2u(n,t)+u(n-1,t))-u(n,t).
	\end{gather*}
	Then, we compute the discrete partial variational derivative (Step 3), which equals zero. Hence (Step 4), we have conservation.
\end{example}
\begin{example}
	\label{heatex1}
	We verify the result from Example \ref{timeexplicitheateqalgo}, using our implementation in {\sc Mathematica} of Algorithm \ref{alg: discreteconservedqexplicit}.

	\begin{small}
		\begin{mmaCell}[moredefined={variables, expression, DiscreteConservedQ}]{Input}
  variables=<|"indVars"\(\pmb{\to}\)\{n\},"depVars"\(\pmb{\to}\)\{u\},
  "rhs"\(\pmb{\to}\)\{u[n]+(u[n+1]-2u[n]+u[n-1])\}|>
  DiscreteConservedQ[variables][u[n]]

@@

\noindent{\bf Step 1: Build the discrete time derivative.}
We build the discrete time derivative by subtracting $\mathcal{J}[u](t)$ from $\mathcal{J}[u](t+1)$. This task is performed in our code using {\sc TimeDifference}.
\begin{example}
	The discrete time derivative for $\sum_n u(n,t)$ is

@@

\noindent {\bf Step 5 to 7: Compute the discrete partial variational derivative and reduce again.}
In the next step, we take the resulting expression and compute its discrete partial variational derivative. Then, we repeat Step 2 to Step 4 applied to the discrete partial variational derivative.
\begin{example}
	\label{heateqalgoimplicit}
	To illustrate Algorithm \ref{alg: discreteconservedqimplicit}, we use the setting of Example \ref{timeexplicitheateqalgo}.
	We compute the discrete time derivative (Step 1)
	\begin{gather*}
		\mathcal{T}=\sum_{n\in\Ii} u(n,t+1)-u(n,t).
	\end{gather*}
	For the translations (Step 2), the stencil of the scheme
	\begin{equation*}
		u(n,t+1)-u(n,t)-(u(n+1,t)-2u(n,t)+u(n-1,t))
	\end{equation*}
	\small
	is
	$
	\Big(\big\{(0,1),(0,0),(1,0),(-1,0)\big\}\Big)
	$
	\normalsize
	with range \small$\Big([-1,1]\times[0,1]\Big)$\normalsize.
	The discrete time derivative has stencil \small$\Big(\big\{(0,1),(0,0)\big\}\Big)$\normalsize 
	and range \small$\Big([0,0]\times[0,1]\Big)$\normalsize.
	The translations are
	\small
	\begin{gather*}
		\Big([0,0]\times[0,1]\Big)-\Big([-1,1]\times[0,1]\Big)=\Big([1,-1]\times[0,0]\Big)=\Big(\emptyset\times[0,0]\Big)
	\end{gather*}
	\normalsize
	which results in no translations, as the range of the stencil of the scheme is greater than the range of the stencil of the discrete time derivative.
	Hence, the Gr{\"o}bner basis (Step 3) coincides with the original scheme for both monomial orders.
	The reduction using the lexicographic monomial order yields as remainder
	\begin{gather*}
		u(n-1,t)-2u(n,t)+u(n+1,t).
	\end{gather*}
	For the reduction using the elimination order, we first eliminate $u(n,t+1)$ and then the remaining instances of $u$.
	The elimination order yields the same remainder as the lexicographic order. Hence, we do not check for the number of instances of time (Step 4).
	We calculate the discrete partial variational derivative (Step 5) of
	\begin{gather*}
		\sum_{n\in\Ii}u(n-1,t)-2u(n,t)+u(n+1,t)
	\end{gather*}
	that equals $0$.
	Hence, repeating the above procedure (Step 6) becomes unnecessary.
	Therefore, (Step 7) we see that the scheme conserves the quantity.
\end{example}
\begin{example}
	We verify our result from Example \ref{heateqalgoimplicit}, using {\sc DiscreteConservedQ}, which detects automatically if the scheme is time-explicit or if
	 the general setting should be applied.

	\begin{small}
		\begin{mmaCell}[moredefined={variables, expression, DiscreteConservedQ}]{Input}
  variables=<|"indVars"\(\pmb{\to}\)\{n,t\},"depVars"\(\pmb{\to}\)\{u\},
  "scheme"\(\pmb{\to}\)\{u[n,t+1]-u[n,t]
  -(u[n+1,t]-2u[n,t]+u[n-1,t])\}|>
  DiscreteConservedQ[variables][u[n,t]]

@@

	\caption{{\sc FindDiscreteConservedQuantityBasis}}
	\label{alg: findconservedquantitybasis}
\end{algorithm}
We have implemented this algorithm in {\sc FindDiscreteConservedQuantityBasis}.
\begin{example}
In the context of the heat equation from Example \ref{heatex1}, we aim to find a basis for conserved quantities generated by $u(n,t)$ and $n$ with a maximum degree of 3.

	\begin{small}
		\begin{mmaCell}[moredefined={variables, FindDiscreteConservedQuantityBasis}]{Input}
  variables=<|"indVars"\(\pmb{\to}\)\{n\},"depVars"\(\pmb{\to}\)\{u\},
  "rhs"\(\pmb{\to}\)\{u[n]+u[n+1]-2u[n]+u[n-1]\}|>
  FindDiscreteConservedQuantityBasis[variables]
  [<|"degree"\(\pmb{\to}\)3,"generators"\(\pmb{\to}\)\{u[n],n\}|>]

@@

The inviscid Burgers equation is the PDE $u_t+uu_x=0$ \cite{smoller94} in one space dimension. Any function of $u$ is a conserved quantity. We search for discretizations preserving $\int u\ dx$.
\begin{example}
	We check for conservation of mass using a forward-difference discretization:

	\begin{small}
		\begin{mmaCell}[moredefined={variables, expression, DiscreteConservedQ}]{Input}
  variables=<|"indVars"\(\pmb{\to}\)\{n,t\},"depVars"\(\pmb{\to}\)\{u\},
  "scheme"\(\pmb{\to}\)\{u[n,t+1]-u[n,t]
  -u[n,t](u[n+1,t]-u[n,t])\}|>
  DiscreteConservedQ[variables][u[n,t]]

\end{mmaCell}
		\begin{mmaCell}{Output}
  False
\end{mmaCell}
	\end{small}
\end{example}
For finding a scheme that preserves mass, we check for conservation under a class of schemes with a parameter.
\begin{example}
	We discretize $u_x$ with a three-point stencil with a parameter $a$.

	\begin{small}
		\begin{mmaCell}[moredefined={variables, expression, DiscreteConservedQ}]{Input}
  variables=<|"indVars"\(\pmb{\to}\)\{n,t\},"depVars"\(\pmb{\to}\)\{u\},
  "pars"\(\pmb{\to}\)\{a\},"scheme"\(\pmb{\to}\)\{u[n,t+1]-u[n,t]
  -u[n,t](a(u[n+1,t]-u[n,t])
  +(1-a)(u[n,t]-u[n-1,t]))\}|>
  DiscreteConservedQ[variables][u[n,t]]

@@

\end{mmaCell}
	\end{small}
The conservation property holds when $a=\frac{1}{2}$, which corresponds to the central difference case.
\end{example}
So far, we have only seen schemes that are in time-explicit form, but our algorithm allows also for general schemes:
\begin{example}
	We consider the scheme from the previous example but in the time-implicit version.

	\begin{small}
		\begin{mmaCell}[moredefined={variables, DiscreteConservedQ}]{Input}
  variables=<|"indVars"\(\pmb{\to}\)\{n,t\},"depVars"\(\pmb{\to}\)\{u\},
  "pars"\(\pmb{\to}\)\{a\},"elimOrder"\(\pmb{\to}\)"implicit",
  "scheme"\(\pmb{\to}\)\{u[n,t+1]-u[n,t]-u[n,t+1]
  (a(u[n+1,t+1]-u[n,t+1])
  +(1-a)(u[n,t+1]-u[n-1,t+1]))\}|>
  DiscreteConservedQ[variables][u[n,t]]

@@

{\bf Conserved quantities for a mean-field game.}
We analyze the conserved quantities preserved by the discretization of a PDE system in the context of mean-field games.
The examples showcase our algorithm's effectiveness in identifying conserved quantities for various discretization schemes, including time-explicit, time-implicit, and mixed schemes.
\begin{example}
	In \cite{gomes18}, Gomes et al. derived the following system
	\begin{gather*}
		v_t+\Big(\frac{v^2}{2}-\frac{m^2}{2}\Big)_x=0,\quad
		m_t-(vm)_x=0.
	\end{gather*}
	This system admits the conserved quantities $\int v\ dx$ and $\int m\ dx$.
	Because this system was derived from a forward-forward mean-field game, we discretize it forward in time. We check with our code if the scheme admits the same preserved quantities as the continuous system:

	\begin{small}
		\begin{mmaCell}[moredefined={variables, DiscreteConservedQ}]{Input}
  variables=<|"indVars"\(\pmb{\to}\)\{n,t\},"depVars"\(\pmb{\to}\)\{v,m\},
  "scheme"\(\pmb{\to}\)\{v[n,t+1]-v[n,t]
  +(\mmaSup{v[n+1,t]}{2}-\mmaSup{m[n+1,t]}{2})/2-(\mmaSup{v[n,t]}{2}-\mmaSup{m[n,t]}{2})/2,
  m[n,t+1]-m[n,t]-m[n+1,t]v[n+1,t]+m[n,t]v[n,t]\}|>
  DiscreteConservedQ[variables][\{v[n,t],m[n,t]\}]

\end{mmaCell}
		\begin{mmaCell}{Output}
  True
\end{mmaCell}
	\end{small}
We replicate this result, observing that the scheme used here is time-explicit:

>>>>

\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage[latin1]{inputenc}
\usepackage{hyperref}
\usepackage{dsfont}
\usepackage[linesnumbered,ruled,vlined]{algorithm2e}
\usepackage{ifthen}
\usepackage{mmacells}
\usepackage[invisible]{MLaTeX}

@@

Here, all quantities, whose conservation in time we check, are functions $\mathcal{J}\in\mathcal{F}(\mathcal{P}(\Ii\times\Nn^l,\Rr^m),\mathcal{D}(\Nn^l))$. 
Let $\tilde{n}:=(n_1,\dots,n_d,s_1,\dots,s_l)\in\Ii\times\Nn^l$. Using the Kronecker delta $\delta$, we rewrite
\begin{gather*}
	\mathcal{J}[u](t_1,\ldots,t_{l})
	=\sum_{\tilde{n}\in\Ii\times\Nn^l}F_{\tilde{n}}[u]\delta_{(t_1,\ldots,t_{l})}(s_1,\ldots,s_l).
\end{gather*}
Let $u,v\in\mathcal{P}(\Ii\times\Nn^l,\Rr^m)$, then
\footnotesize
\begin{gather*}
	D^l\mathcal{J}[u](v)(t_1,\ldots,t_{l}):=\frac{d}{d\epsilon}\mathcal{J}[u+\epsilon v]\Big|_{\epsilon=0}(t_1,\ldots,t_{l})\\
	=\sum_{\tilde{n}}D_1G(u(\tilde{n}+e^1),\ldots,u(\tilde{n}+e^k))\delta_{(t_1,\ldots,t_{l})}(s_1,\ldots,s_l)v(\tilde{n}+e^1)\\
	+D_2G(u(\tilde{n}+e^1),\ldots,u(\tilde{n}+e^k))\delta_{(t_1,\ldots,t_{l})}(s_1,\ldots,s_l)v(\tilde{n}+e^2)\\+\ldots
	+D_kG(u(\tilde{n}+e^1),\ldots,u(\tilde{n}+e^k))\delta_{(t_1,\ldots,t_{l})}(s_1,\ldots,s_l)v(\tilde{n}+e^k)\\
	=\sum_{\tilde{n}}D_1G(u(\tilde{n}),\ldots,u(\tilde{n}+e^k-e^1))\delta_{(t_1,\ldots,t_{l})}(s_1-e^1_{d+1},\ldots,s_l-e^1_{d+l})v(\tilde{n})\\
	+D_2G(u(\tilde{n}+e^1-e^2),\ldots,u(\tilde{n}+e^k-e^2))\delta_{(t_1,\ldots,t_{l})}(s_1-e^2_{d+1},\ldots,s_l-e^2_{d+l})v(\tilde{n})\\+\ldots
	+D_kG(u(\tilde{n}+e^1-e^k),\ldots,u(\tilde{n}))\delta_{(t_1,\ldots,t_{l})}(s_1-e^k_{d+1},\ldots,s_l-e^k_{d+l})v(\tilde{n})\\
	=\sum_{\tilde{n}}\Big(D_1G(u(\tilde{n}),\ldots,u(\tilde{n}+e^k-e^1))\delta_{(t_1+e^1_{d+1},\ldots,t_{l}+e^1_{d+l})}(s_1,\ldots,s_l)\\
	+D_2G(u(\tilde{n}+e^1-e^2),\ldots,u(\tilde{n}+e^k-e^2))\delta_{(t_1+e^2_{d+1},\ldots,t_{l}+e^2_{d+l})}(s_1,\ldots,s_l)\\+\ldots
	+D_kG(u(\tilde{n}+e^1-e^k),\ldots,u(\tilde{n}))\delta_{(t_1+e^k_{d+1},\ldots,t_{l}+e^k_{d+l})}(s_1,\ldots,s_l)\Big)v(\tilde{n})\\
	=:\sum_{\tilde{n}}\mathcal{V}^l(\mathcal{J})[u](\tilde{n},t)v(\tilde{n}).
\end{gather*}
\normalsize
\begin{definition}
$\mathcal{V}^l(\mathcal{J})$ is the \emph{discrete partial variational derivative} of $\mathcal{J}$.
	Note that $\mathcal{V}^0=\mathcal{V}$.
\end{definition}
\begin{example}
	\label{dpvd1}
	The discrete partial variational derivative of 
	\begin{gather*}
		\mathcal{J}[u](t)=\sum_nu(n,t+1)-u(n,t)=\sum_{n,s}(u(n,s+1)-u(n,s))\delta_t(s)
	\end{gather*}
	is $\delta_{t+1}(s)-\delta_t(s)$, because
	\small
	\begin{gather*}
		D^1(\mathcal{J})[u](v)(t)\\
		=\frac{d}{d\epsilon} \sum_{n,s}((u(n,s+1)+\epsilon v(n,s+1))-(u(n,s)+\epsilon v(n,s)))\delta_t(s)\\
		=\sum_{n,s}v(n,s+1)\delta_t(s)-v(n,s)\delta_t(s)=\sum_{n,s}v(n,s)\delta_t(s-1)-v(n,s)\delta_t(s)\\
		=\sum_{n,s}(\delta_{t+1}(s)-\delta_t(s))v(n,s)=\sum_{n,s}\mathcal{V}^1(\mathcal{J})(n,s,t)v(n,s).
	\end{gather*}
	\normalsize
\end{example}
\begin{example}
	In our {\sc Mathematica} implementation, Example \ref{dpvd1} is computed by
	\begin{EXE}
		LABEL="dpvd1"
		(*CODE*)
		variables = Association["indVars"->{n,t},"depVars"->{u},"timeVars"->{t}]
		expression = u[n,t+1] - u[n,t]
		PartialDVarDOperator[variables][expression]
		(*CODE*)
		PUNCT={,.}
	\end{EXE}	
	\begin{small}
		\begin{mmaCell}[moredefined={variables, expression, PartialDVarD}]{Input}
  variables=<|"indVars"\(\pmb{\to}\)\{n,t\},"depVars"\(\pmb{\to}\)\{u\},
  "timeVars"\(\pmb{\to}\)\{t\}|>
  PartialDVarD[variables][u[n,t+1]-u[n,t]]

@@

	\caption{{\sc time-explicit DiscreteConservedQ}}
	\label{alg: discreteconservedqexplicit}
\end{algorithm}
The result of Step 2 exists only at time $t$ because all instances of $t+1$ have been replaced. Thus, we do not need any computations of difference ideals or Gr{\"o}bner basis.
\begin{example}
	\label{timeexplicitheateqalgo}
	We examine the conservation of $\sum_nu(n,t)$ under the following scheme.
	\begin{equation*}
		u(n,t+1)-u(n,t)-(u(n+1,t)-2u(n,t)+u(n-1,t)).
	\end{equation*}
	The discrete time derivative (Step 1) is
	\begin{gather*}
		\sum_{n\in\Ii}u(n,t+1)-u(n,t).
	\end{gather*}
	Replace $u(n,t+1)$ by the right-hand side (Step 2) to get
	\begin{gather*}
		\sum_{n\in\Ii}u(n,t)+(u(n+1,t)-2u(n,t)+u(n-1,t))-u(n,t).
	\end{gather*}
	Then, we compute the discrete partial variational derivative (Step 3), which equals zero. Hence (Step 4), we have conservation.
\end{example}
\begin{example}
	\label{heatex1}
	We verify the result from Example \ref{timeexplicitheateqalgo}, using our implementation in {\sc Mathematica} of Algorithm \ref{alg: discreteconservedqexplicit}.
	\begin{EXE}
		LABEL="rhs1"
		(*CODE*)
		variables = Association["indVars"->{n},"depVars"->{u},"rhs"->{u[n]+(u[n+1]-2*u[n]+u[n-1])}]
		expression = u[n]
		DiscreteConservedQOperator[variables][expression]
		(*CODE*)
		PUNCT={,.}
	\end{EXE}
	\begin{small}
		\begin{mmaCell}[moredefined={variables, expression, DiscreteConservedQ}]{Input}
  variables=<|"indVars"\(\pmb{\to}\)\{n\},"depVars"\(\pmb{\to}\)\{u\},
  "rhs"\(\pmb{\to}\)\{u[n]+(u[n+1]-2u[n]+u[n-1])\}|>
  DiscreteConservedQ[variables][u[n]]

@@

\noindent{\bf Step 1: Build the discrete time derivative.}
We build the discrete time derivative by subtracting $\mathcal{J}[u](t)$ from $\mathcal{J}[u](t+1)$. This task is performed in our code using {\sc TimeDifference}.
\begin{example}
	The discrete time derivative for $\sum_n u(n,t)$ is
	\begin{EXE}
		LABEL="heateq2"
		(*CODE*)
		variables = <|"indVars"->{n,t},"depVars"->{u}|>
		expression = u[n,t]
		TimeDifferenceOperator[variables][expression]
		(*CODE*)
		PUNCT={,.}
	\end{EXE}

@@

\noindent {\bf Step 5 to 7: Compute the discrete partial variational derivative and reduce again.}
In the next step, we take the resulting expression and compute its discrete partial variational derivative. Then, we repeat Step 2 to Step 4 applied to the discrete partial variational derivative.
\begin{example}
	\label{heateqalgoimplicit}
	To illustrate Algorithm \ref{alg: discreteconservedqimplicit}, we use the setting of Example \ref{timeexplicitheateqalgo}.
	We compute the discrete time derivative (Step 1)
	\begin{gather*}
		\mathcal{T}=\sum_{n\in\Ii} u(n,t+1)-u(n,t).
	\end{gather*}
	For the translations (Step 2), the stencil of the scheme
	\begin{equation*}
		u(n,t+1)-u(n,t)-(u(n+1,t)-2u(n,t)+u(n-1,t))
	\end{equation*}
	\small
	is
	$
	\Big(\big\{(0,1),(0,0),(1,0),(-1,0)\big\}\Big)
	$
	\normalsize
	with range \small$\Big([-1,1]\times[0,1]\Big)$\normalsize.
	The discrete time derivative has stencil \small$\Big(\big\{(0,1),(0,0)\big\}\Big)$\normalsize 
	and range \small$\Big([0,0]\times[0,1]\Big)$\normalsize.
	The translations are
	\small
	\begin{gather*}
		\Big([0,0]\times[0,1]\Big)-\Big([-1,1]\times[0,1]\Big)=\Big([1,-1]\times[0,0]\Big)=\Big(\emptyset\times[0,0]\Big)
	\end{gather*}
	\normalsize
	which results in no translations, as the range of the stencil of the scheme is greater than the range of the stencil of the discrete time derivative.
	Hence, the Gr{\"o}bner basis (Step 3) coincides with the original scheme for both monomial orders.
	The reduction using the lexicographic monomial order yields as remainder
	\begin{gather*}
		u(n-1,t)-2u(n,t)+u(n+1,t).
	\end{gather*}
	For the reduction using the elimination order, we first eliminate $u(n,t+1)$ and then the remaining instances of $u$.
	The elimination order yields the same remainder as the lexicographic order. Hence, we do not check for the number of instances of time (Step 4).
	We calculate the discrete partial variational derivative (Step 5) of
	\begin{gather*}
		\sum_{n\in\Ii}u(n-1,t)-2u(n,t)+u(n+1,t)
	\end{gather*}
	that equals $0$.
	Hence, repeating the above procedure (Step 6) becomes unnecessary.
	Therefore, (Step 7) we see that the scheme conserves the quantity.
\end{example}
\begin{example}
	We verify our result from Example \ref{heateqalgoimplicit}, using {\sc DiscreteConservedQ}, which detects automatically if the scheme is time-explicit or if
	 the general setting should be applied.
	\begin{EXE}
		LABEL="heateq10"
		(*CODE*)
		variables = Association["indVars"->{n,t},"depVars"->{u},"scheme"->{u[n,t+1]-u[n,t]-(u[n+1,t]-2*u[n,t]+u[n-1,t])}]
		expression = u[n,t]
		DiscreteConservedQOperator[variables][expression]
		(*CODE*)
		PUNCT={,.}
	\end{EXE}
	\begin{small}
		\begin{mmaCell}[moredefined={variables, expression, DiscreteConservedQ}]{Input}
  variables=<|"indVars"\(\pmb{\to}\)\{n,t\},"depVars"\(\pmb{\to}\)\{u\},
  "scheme"\(\pmb{\to}\)\{u[n,t+1]-u[n,t]
  -(u[n+1,t]-2u[n,t]+u[n-1,t])\}|>
  DiscreteConservedQ[variables][u[n,t]]

@@

	\caption{{\sc FindDiscreteConservedQuantityBasis}}
	\label{alg: findconservedquantitybasis}
\end{algorithm}
We have implemented this algorithm in {\sc FindDiscreteConservedQuantityBasis}.
\begin{example}
In the context of the heat equation from Example \ref{heatex1}, we aim to find a basis for conserved quantities generated by $u(n,t)$ and $n$ with a maximum degree of 3.
	\begin{EXE}
		LABEL="heateq11"
		(*CODE*)
		variables = <|"indVars"->{n},"depVars"->{u},"rhs"->{u[n]+u[n+1]-2*u[n]+u[n-1]}|>
		FindDiscreteConservedQuantityBasisOperator[variables][<|"degree"->3,"generators"->{u[n],n}|>]
		(*CODE*)
		PUNCT={,.}
	\end{EXE}
	\begin{small}
		\begin{mmaCell}[moredefined={variables, FindDiscreteConservedQuantityBasis}]{Input}
  variables=<|"indVars"\(\pmb{\to}\)\{n\},"depVars"\(\pmb{\to}\)\{u\},
  "rhs"\(\pmb{\to}\)\{u[n]+u[n+1]-2u[n]+u[n-1]\}|>
  FindDiscreteConservedQuantityBasis[variables]
  [<|"degree"\(\pmb{\to}\)3,"generators"\(\pmb{\to}\)\{u[n],n\}|>]

@@

The inviscid Burgers equation is the PDE $u_t+uu_x=0$ \cite{smoller94} in one space dimension. Any function of $u$ is a conserved quantity. We search for discretizations preserving $\int u\ dx$.
\begin{example}
	We check for conservation of mass using a forward-difference discretization:
	\begin{EXE}
		LABEL="burgers1"
		(*CODE*)
		variables = Association["indVars"->{n,t},"depVars"->{u},"scheme"->{u[n,t+1]-u[n,t]-u[n,t]*(u[n+1,t]-u[n,t])}];
		expression = u[n,t];
		DiscreteConservedQOperator[variables][expression]
		(*CODE*)
		PUNCT={,.}
	\end{EXE}
	\begin{small}
		\begin{mmaCell}[moredefined={variables, expression, DiscreteConservedQ}]{Input}
  variables=<|"indVars"\(\pmb{\to}\)\{n,t\},"depVars"\(\pmb{\to}\)\{u\},
  "scheme"\(\pmb{\to}\)\{u[n,t+1]-u[n,t]
  -u[n,t](u[n+1,t]-u[n,t])\}|>
  DiscreteConservedQ[variables][u[n,t]]

\end{mmaCell}
		\begin{mmaCell}{Output}
  False
\end{mmaCell}
	\end{small}
\end{example}
For finding a scheme that preserves mass, we check for conservation under a class of schemes with a parameter.
\begin{example}
	We discretize $u_x$ with a three-point stencil with a parameter $a$.
	\begin{EXE}
		LABEL="burgers2"
		(*CODE*)
		variables = Association["indVars"->{n,t}, "depVars"->{u}, "pars"->{a}, "scheme"->{u[n,t+1]-u[n,t]-u[n,t]*(a*(u[n+1,t]-u[n,t])+(1-a)*(u[n,t]-u[n-1,t]))}];
		expression = u[n,t];
		DiscreteConservedQOperator[variables][expression]
		(*CODE*)
		PUNCT={,.}
	\end{EXE}
	\begin{small}
		\begin{mmaCell}[moredefined={variables, expression, DiscreteConservedQ}]{Input}
  variables=<|"indVars"\(\pmb{\to}\)\{n,t\},"depVars"\(\pmb{\to}\)\{u\},
  "pars"\(\pmb{\to}\)\{a\},"scheme"\(\pmb{\to}\)\{u[n,t+1]-u[n,t]
  -u[n,t](a(u[n+1,t]-u[n,t])
  +(1-a)(u[n,t]-u[n-1,t]))\}|>
  DiscreteConservedQ[variables][u[n,t]]

@@

\end{mmaCell}
	\end{small}
The conservation property holds when $a=\frac{1}{2}$, which corresponds to the central difference case.
\end{example}
So far, we have only seen schemes that are in time-explicit form, but our algorithm allows also for general schemes:
\begin{example}
	We consider the scheme from the previous example but in the time-implicit version.
	\begin{EXE}
		LABEL="burgers3"
		(*CODE*)
		variables = Association["indVars"->{n,t}, "depVars"->{u}, "pars"->{a},"elimOrder"->"implicit", "scheme"->{u[n,t+1]-u[n,t]-u[n,t+1]*(a*(u[n+1,t+1]-u[n,t+1])+(1-a)*(u[n,t+1]-u[n-1,t+1]))}]
		DiscreteConservedQOperator[variables][u[n,t]]
		(*CODE*)
		PUNCT={,.}
	\end{EXE}
	\begin{small}
		\begin{mmaCell}[moredefined={variables, DiscreteConservedQ}]{Input}
  variables=<|"indVars"\(\pmb{\to}\)\{n,t\},"depVars"\(\pmb{\to}\)\{u\},
  "pars"\(\pmb{\to}\)\{a\},"elimOrder"\(\pmb{\to}\)"implicit",
  "scheme"\(\pmb{\to}\)\{u[n,t+1]-u[n,t]-u[n,t+1]
  (a(u[n+1,t+1]-u[n,t+1])
  +(1-a)(u[n,t+1]-u[n-1,t+1]))\}|>
  DiscreteConservedQ[variables][u[n,t]]

@@

{\bf Conserved quantities for a mean-field game.}
We analyze the conserved quantities preserved by the discretization of a PDE system in the context of mean-field games.
The examples showcase our algorithm's effectiveness in identifying conserved quantities for various discretization schemes, including time-explicit, time-implicit, and mixed schemes.
\begin{example}
	In \cite{gomes18}, Gomes et al. derived the following system
	\begin{gather*}
		v_t+\Big(\frac{v^2}{2}-\frac{m^2}{2}\Big)_x=0,\quad
		m_t-(vm)_x=0.
	\end{gather*}
	This system admits the conserved quantities $\int v\ dx$ and $\int m\ dx$.
	Because this system was derived from a forward-forward mean-field game, we discretize it forward in time. We check with our code if the scheme admits the same preserved quantities as the continuous system:
	\begin{EXE}
		LABEL="mfg1"
		(*CODE*)
		variables = Association[
		"indVars" -> {n, t},
		"depVars" -> {v, m},
		"display result" -> True,
		"scheme" -> {v[n, t + 1] - 
			v[n, 
			t] + (v[n + 1, t]^2/2 - m[n + 1, t]^2/2) - (v[n, t]^2/2 - 
			m[n, t]^2/2), 
			m[n, t + 1] - 
			m[n, t] - (m[n + 1, t]*v[n + 1, t]) + (m[n, t]*v[n, t])}
		];
		DiscreteConservedQOperator[variables][{v[n, t], m[n, t]}]
		(*CODE*)
		PUNCT={,.}
	\end{EXE}
	\begin{small}
		\begin{mmaCell}[moredefined={variables, DiscreteConservedQ}]{Input}
  variables=<|"indVars"\(\pmb{\to}\)\{n,t\},"depVars"\(\pmb{\to}\)\{v,m\},
  "scheme"\(\pmb{\to}\)\{v[n,t+1]-v[n,t]
  +(\mmaSup{v[n+1,t]}{2}-\mmaSup{m[n+1,t]}{2})/2-(\mmaSup{v[n,t]}{2}-\mmaSup{m[n,t]}{2})/2,
  m[n,t+1]-m[n,t]-m[n+1,t]v[n+1,t]+m[n,t]v[n,t]\}|>
  DiscreteConservedQ[variables][\{v[n,t],m[n,t]\}]
